---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dendro

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/dendro)](https://CRAN.R-project.org/package=dendro)
[![Codecov test coverage](https://codecov.io/gh/mikemahoney218/dendro/branch/main/graph/badge.svg)](https://app.codecov.io/gh/mikemahoney218/dendro?branch=main)
[![R-CMD-check](https://github.com/mikemahoney218/dendro/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikemahoney218/dendro/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

dendro is an experimental package I wrote to play around with tree structures and the new (as of this writing) [S7](https://github.com/RConsortium/OOP-WG/) class system. It's not intended for serious workloads, but rather as a way for me to kick the wheels on S7 and toy with recursive operations.

## Installation

You can install the development version of dendro like so:

``` r
# install.packages("remotes")
remotes::install_github("RConsortium/OOP-WG")
remotes::install_github("mikemahoney218/dendro")
```

## Example

The core of the dendro package lives in two S7 classes, `node()` and `tree()`. The node class has four slots: the index value of the node within the tree, the index of the node's parent, and the index of any children the node has, as well as a list slot for any attributes you want to attach to the node:

```{r example}
library(dendro)

node(
  index = 1L,
  value = list(), 
  parent = 0L,
  children = 2L
)
```

Nodes are almost entirely unvalidated -- there's no requirement a parent knows about all its children, that a node's parent exists, that the node's index is unique, or that its index corresponds to its location in a tree. Node-creating code is responsible for those validations. 

The tree class is slightly more interesting. Its constructor function takes a list of nodes:

```{r}
tree(list(node(1L)))
```

The tree class also doesn't do a ton of validation; it only checks to make sure that every node (save for one, the root node) has a parent. However, this class also comes with a few useful dynamic slots: `size`, the number of nodes in the tree; `node_indices`, a vector of all `index` values in the nodes list; `next_idx`, a safe position for the next node to be inserted at, and `root`, the index of the root node in the nodes list.

There's also a handful of methods defined for tree classes. For instance, `add_node()` will automatically add a node at the next available index, for programmatic tree construction:

```{r}
example_tree <- tree() |>
  add_node(name = "root") |>
  add_node(list(), parent = 1L) |>
  add_node(list(), parent = 1L) |>
  add_node(list(), parent = 2L) |>
  add_node(list(), parent = 4L)

example_tree
```

Note that `add_node()` is relatively slow, as the tree is re-validated after each call to `add_node()`. To make an entirely new tree from scratch, it's usually faster to call `tree()` with a list of nodes instead.

You can call `drop_node()` to drop nodes from the tree. Note that all descendants of the node will be dropped as well, as dendro doesn't allow multiple roots in a tree:

```{r}
example_tree |> 
  drop_node(1)
```

```{r}
# Get the depth of a node in the tree
example_tree |> 
  node_depth(5)

# Get the distance between two nodes in the tree
example_tree |> 
  distance_to_target(1, 5)

# Get the indices of the children for a node
example_tree |> 
  get_children(1)

# Get the indices of _all_ descendants for a node
example_tree |> 
  get_all_descendants(1)

# Get node objects from a tree
example_tree |> 
  get_nodes(1)

# Get the value slot of a node object from a tree
example_tree |> 
  get_node_values(1)
```

There's also a function, `as_tree`, for converting data frames to tree formats:

```{r}
data.frame(x = letters[1:4], y = letters[25:26]) |> 
  as_tree(c("x", "y"), root_name = "root")

data.frame(x = letters[1:4], y = letters[25:26]) |> 
  as_tree(c("y", "x"), root_name = "root")
```

